something about variable-byte encoding

### 2020-02-12 gradu/msc_english_contents.tex
DAC with select

Rank and select are two succinct data structure operations which operate on a bit array B. $Rank_1$(B, i) gives the sum of 1 bits between B[0] and B[i] and 
$select_1$(b, i) gives the index of i-th 1 bit in B. Both operations work in constant time (citation?) and they require only a few percents of extra 
space over the bit array B (citation?). Locations of 1 bits in B should represent locations of the items in the encoded data. For most compression algorithms, 
this requires B to be created in addition to the existing data and the length of B usually has to be close to the length of the data.

VB encoding has several advantages with search and rank: it's data is compressed in blocks of equal length which significantly shrinks the length of B. Also 
the bit array C formed from the continuation bits already stores the locations of items. In this case, $rank_1$(C, i) would give the sum of end bytes up to 
i-th index and $select_1$(C, i) would give the location of the ending byte of i-th compressed element.

@@

(TODO: check if actually first introduced or just first with rank)

do rank/select need to be explained better? (or just referenced to)

@@

- TODO: find out where rank+select initially come from.

 An effective
version of random access has already been introduced \citep{Bri09}.

@@

Using $select_1$ on the continuation bit array to achieve direct access is more intuitive than using $rank_1$. The element locations are already marked with 1's 
in $B$ and a single $select_1$ query gets the desired starting point.

\chapter{Experimental results}
 - comparison to basic implementation + Bri09

 - do we need to support search()?

@@

\section{Figures}
Figure gives an example how to add figures to the document. Remember always to cite the figure in the main text.

\section{Tables}

Table gives an example how to report experimental results. Remember always to cite the table in the main text.

>>>>

Rank and select are two succinct data structure operations which operate on a bit array B. $Rank_1$(B, i) gives the sum of 1 bits between B[0] and B[i] and 
$select_1$(b, i) gives the index of i-th 1 bit in B. Both operations work in constant time (citation?) and they require only a few percents of extra 
space over the bit array B. See Table~\ref{supportsize} for $rank$ and $select$ size requirement approximations on different sizes of bit arrays.

Locations of 1 bits in B should represent locations of the items in the encoded data. For most compression algorithms, this requires B to be created in addition 
to the existing data and the length of B usually has to be close to the length of the data. VB encoding has several advantages with search and rank: it's data 
is compressed in blocks of equal length which significantly shrinks the length of B. Also the bit array C formed from the continuation bits already stores the 
locations of items. In this case, $rank_1$(C, i) would give the sum of end bytes up to i-th index and $select_1$(C, i) would give the location of the ending 
byte of i-th compressed element.

\begin{table}
\centering
\caption{Size of support structure with SDSL-lite library}
\begin{tabular}{l||c c c c c c} 
Structure & TODO & get & multiple & size & calculations & here\\ 
\hline \hline 
$rank$ & 34.97 & 49 & 53.04 & 52.18 & 53.08 & 76.21\\
$select$ & 33.57 & 32.47 & 42.96 & 43.11 & 46.15 & 65.14\\
\hline
%
\end{tabular}
\label{supportsize}
\end{table}

@@

do rank/select algorithms + data structures need to be explained better? (or just referenced to)

@@

@@

Using $select_1$ on the continuation bit array to achieve direct access is more intuitive than using $rank_1$. The element locations are already marked with 1's 
in $B$ and a single $select_1$ query gets the desired starting point, while the forementioned version \citep{Bri09} used one $rank$ query for each chunk beyond 
the first. Minimizing the amount of $select$ and $rank$ queries is important. They run in constant time but their impact is huge, because rest of the VB decoding 
consists of a few bit operations.

To use $select_1$ with VB, continuation bits need to be separated from chunks to their own bit array and a $select_1$ structure built over it. $Select$(i) returns the
location of the end byte of i-th element. Therefore the start of j-th element in the block array is at block $b_s$ = $select_1$(j-1) + 1. To simplify calculations even further,
block sizes of 2, 4 and 8 were used to avoid splitting of blocks between bytes.

The blocks should be stored in an order that, depending on the endianness of the system, when reading an unsigned integer from the block array, the blocks are already 
in correct order. This way two bit shift operations are enough to extract the desired value.  Assume original element size is 32 bits, block length is b and requested
element is at position x in the block array and it's compressed within k blocks. Starting byte s of element e is x * b div 8 where div is the integer division. Offset o
is the remainder of previous division, o = x*b mod 8. 32-bit value e is read from memory from s. Then we bit shift left 32-b*k-o to remove trailing bits and bit shift right
32-b*k to re-align bits and remove leading bits.

Simple way to calculate block length k of i-th element is to get $select_1$(i+1) and subtract already calculated start block index from it. This causes a second $select_1$
query, which is suboptimal. A much faster option is to iterate forward in bit array until next 1 is found. An even faster option is to read an integer from the bit array, fix
it's offset and count trailing zeros. Traveling the bit array is recommended, since it is very simple to implement and is not noticably slower. See Table~\ref{endblockcall} for 
detailed times.

\begin{table}
\centering
\caption{Time to calculate end blocks}
\begin{tabular}{l||c c c c c c} 
Version & Average time over 10Mall \\ 
\hline \hline 
select(i+1) & 34.97 \\
iterate bit array & 34.97 \\
bit shift magic & 34.97 \\
\hline
%
\end{tabular}
\label{endblockcall}
\end{table}

(maybe to future work) Storing bytes in this fashion may have benefits when reading subsequent (fast next() and previous()?) elements. With some modifications, SIMD instructions can possibly be applied.

TODO: check if rank query takes less when done on single byte array instead of multiples

 - get actual results of rank vs select support size
 - take advantage of how data is stored

\chapter{Experimental results}
 - data sets

- comparison to basic implementation + Bri09
  - compare with b=2,4,8

 - do we need to support search()?
 - SIMD?

@@
